Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations

### 2023-08-15 index.qmd
moved to quarto

![The embedding process (**A**) can help to identify links (interactions) that
may have been missed within the original community (represented by the orange
dashed arrows, **B**). Transfer learning (**D**) allows for the prediction links
(interactions) even when novel species (**C**) are included alongside the
original community. This is achieved by learning using other relevant predictors
(*e.g.* traits) in conjunction with the known interactions to infer latent
values (**E**). Ultimately this allows us to predict links (interactions) for
species external from the original sample (blue dashed arrows) as well as
missing within sample links (**F**). Within this context the predicted (and
original) networks as well as the ecological predictors used (green boxes) are
products that can be quantified through measurements in the field, whereas the
embedded as well as imputed matrices (purple box) are representative of a
decomposition of the interaction matrices onto the embedding
space](figures/conceptual_2.png){#fig:embedding}

@@

Graph (or network) embedding (@fig:embedding) is a family of machine learning
techniques, whose main task is to learn a mapping function from a discrete graph
to a continuous domain [@Arsov2019Network; @Chami2022Machine]. Their main goal
is to learn a low dimensional vector representation of the graph (embeddings),
such that its key properties (*e.g.* local or global structures) are retained in
the embedding space [@Yan2005Graph]. The embedding space may, but will not
necessarily, have lower dimensionality than the graph. Ecological networks are
promising candidates for the routine application of embeddings, as they tend to
possess a shared structural backbone [see *e.g.* @BramonMora2018Identifying],
which hints at structural invariants in empirical data. Assuming that these
structural invariants are common enough, they would dominate the structure of
networks, and therefore be adequately captured by the first (lower) dimensions
of an embedding, without the need to measure derived aspects of their structure
(*e.g.* motifs, paths, modularity, ...).

@@

Food webs are inherently low-dimensional objects, and can be adequately
represented with less than ten dimensions [@Braga2021Phylogenetic;
@Eklof2013Dimensionality; @Braga2019Spatial]. Simulation results by
@Botella2022Appraisal suggested that there is no dominant method to identify
architectural similarities between networks: multiple approaches need to be
tested and compared to the network descriptor of interest on a problem-specific
basis. This matches previous results on graph embedding, wherein different
embedding algorithms yield different network embeddings [@Goyal2018Graph],
calling for a careful selection of the problem-specific approach to use. In
@tbl:methods, we present a selection of common graph and node embedding methods,
alongside examples of their use to predict interactions or statistical
associations between species. These methods rely largely on linear algebra or
pseudo-random walks on graphs. All forms of embeddings presented in @tbl:methods
share the common property of summarizing their objects into (sets of) dense
feature vectors, that capture the overall network structure, pairwise
information on nodes, and emergent aspects of the network, in a compressed way
(*i.e.* with some information loss, as we later discuss in the illustration).
Node embeddings tend to focus on maintaining pairwise relationships (*i.e.*
species interactions), while graph embeddings focus on maintaining the network
structure (*i.e.* emergent properties). Nevertheless, some graph embedding
techniques [like RDPG, see *e.g.* @Wu2021Maximum] will provide high-quality
node-level embeddings while also preserving network structure.

::: {#box-gnn .callout-note}
## Graph Neural Networks {#box-gnn}

One prominent family of approaches we do not discuss in the present manuscript
is Graph Neural Networks [GNN; @Zhou2020Graph]. GNN are, in a sense, a method to
embed a graph into a dense subspace, but belong to the family of deep learning
methods, which has its own set of practices [see *e.g.* @Goodfellow2016Deep]. An
important issue with methods based on deep learning is that, because their
parameter space is immense, the sample size of the data fed into them must be
similarly large (typically thousands of instances). This is a requirement for
the model to converge correctly during training, but this assumption is unlikely
to be met given the size of datasets currently available for metawebs (or single
time/location species interaction networks). This data volume requirement is
mostly absent from the techniques we list below. Furthermore, GNN still have
some challenges related to their shallow structure, and concerns related to
scalability [see @Gupta2021Graph for a review], which are mostly absent from the
methods listed in @tbl:methods. Assuming that the uptake of next-generation
biomonitoring techniques does indeed deliver larger datasets on species
interactions [@Bohan2017Nextgeneration], there is nevertheless the potential for
GNN to become an applicable embedding/predictive technique in the coming years.
:::

Graph embeddings *can* serve as a dimensionality reduction method. For example,
RDPG [@Strydom2022Food] and t-SVD [truncated Singular Value Decomposition;
@Poisot2021Imputing] typically embed networks using fewer dimensions than the
original network [the original network has as many dimensions as species, and as
many informative dimensions as trophically unique species; @Strydom2021Svd].
However, this is not necessarily the case -- indeed, one may perform a PCA (a
special case of SVD) to project the raw data into a subspace that improves the
efficacy of t-SNE [t-distributed stochastic neighbor embedding;
@Maaten2009Learning]. There are many dimensionality reductions
[@Anowar2021Conceptual] that can be applied to an embedded network should the
need for dimensionality reduction (for example for data visualization) arise. In
brief, many graph embeddings *can* serve as dimensionality reduction steps, but
not all do, neither do all dimensionality reduction methods provide adequate
graph embedding capacities. In the next section (and @fig:embedding), we show
how the amount of dimensionality reduction can affect the quality of the
embedding.

| Method        | Object          | Technique                        | Reference                | Application                                                                                                       |
| ------------- | --------------- | -------------------------------- | ------------------------ | ----------------------------------------------------------------------------------------------------------------- |
| tSNE          | nodes           | statistical divergence           | @Hinton2002Stochastic    | [@Cieslak2020Tdistributed, species-environment responses $^a$] [@Gibb2021Data, host-virus network representation] |
| LINE          | nodes           | stochastic gradient descent      | @Tang2015Line            |                                                                                                                   |
| SDNE          | nodes           | gradient descent                 | @Wang2016Structural      |                                                                                                                   |
| node2vec      | nodes           | stochastic gradient descent      | @Grover2016Node2vec      |                                                                                                                   |
| HARP          | nodes           | meta-strategy                    | @Chen2017Harp            |                                                                                                                   |
| DMSE          | joint nodes     | deep neural network              | @Chen2017Deep            | [@Chen2017Deep, species-environment interactions $^b$]                                                            |
| graph2vec     | sub-graph       | skipgram network                 | @Narayanan2017Graph2vec  |                                                                                                                   |
| RDPG          | graph           | SVD                              | @Young2007Random         | [@DallaRiva2016Exploring, trophic interactions] [@Poisot2021Imputing, host-virus network prediction]              |
| GLEE          | graph           | Laplacian eigenmap               | @Torres2020Glee          |                                                                                                                   |
| DeepWalk      | graph           | stochastic gradient descent      | @Perozzi2014Deepwalk     | [@Wardeh2021Predicting, host-virus interactions]                                                                  |
| GraphKKE      | graph           | stochastic differential equation | @Melnyk2020Graphkke      | [@Melnyk2020Graphkke, microbiome species associations $^a$]                                                       |
| FastEmbed     | graph           | eigen decomposition              | @Ramasamy2015Compressive |                                                                                                                   |
| PCA           | graph           | eigen decomposition              | @Surendran2013Graph      | [@Strydom2021Roadmap, host-parasite interactions]                                                                 |
| Joint methods | multiple graphs | multiple strategies              | @Wang2021Joint           |                                                                                                                   |

: Overview of some common graph embedding approaches, by type of embedded
objects, alongside examples of their use in the prediction of species
interactions. These methods have not yet been routinely used to predict species
interactions; most examples that we identified were either statistical
associations, or analogues to joint species distribution models. $^a$:
application is concerned with *statistical* interactions, which are not
necessarilly direct biotic interactions; $^b$:application is concerned with
joint-SDM-like approach, which is also very close to statistical associations as
opposed to direct biotic interactions. Given the need to evaluate different
methods on a problem-specific basis, the fact that a lot of methods have not
been used on network problems is an opportunity for benchmarking and method
development. Note that the row for PCA also applies to kernel/probabilistic PCA,
which are variations on the more general method of SVD. Note further that tSNE
has been included because it is frequently used to embed graphs, including of
species associations/interactions, despite not being strictly speaking, a graph
embedding technique [see *e.g.* @Chami2022Machine]. {#tbl:methods}

The popularity of graph embedding techniques in machine learning is more than
the search for structural invariants: graphs are discrete objects, and machine
learning techniques tend to handle continuous data better. Bringing a sparse
graph into a continuous, dense vector space [@Xu2021Understanding] opens up a
broader variety of predictive algorithms, notably of the sort that are able to
predict events as probabilities [@Murphy2022Probabilistic]. Furthermore, the
projection of the graph itself is a representation that can be learned;
@Runghen2021Exploiting, for example, used a neural network to learn the
embedding of a network in which not all interactions were known, based on the
nodes' metadata. This example has many parallels in ecology (see @fig:embedding
**C**), in which node metadata can be represented by phylogeny, abundance, or
functional traits. Using phylogeny as a source of information assumes (or
strives to capture) the action of evolutionary processes on network structure,
which at least for food webs have been well documented [@Braga2021Phylogenetic;
@DallaRiva2016Exploring; @Eklof2016Phylogenetic; @Stouffer2007Evidence;
@Stouffer2012Evolutionary]; similarly, the use of functional traits assumes that
interactions can be inferred from the knowledge of trait-matching rules, which
is similarly well supported in the empirical literature
[@Bartomeus2013Understanding; @Bartomeus2016Common; @Goebel2023Body;
@Gravel2013Inferring]. Relating this information to an embedding rather than a
list of network measures would allow to capture their effect on the more
fundamental aspects of network structure; conversely, the absence of a
phylogenetic or functional signal may suggest that evolutionary/trait processes
are not strong drivers of network structure, therefore opening a new way to
perform hypothesis testing.

@@

![Validation of an embedding for a host-parasite metaweb, using Random Dot
Product Graphs. **A**, decrease in approximation error as the number of
dimensions in the subspaces increases. **B**, increase in cumulative variance
explained as the number of ranks considered increases; in **A** and **B**, the
dot represents the point of inflexion in the curve (at rank 39) estimated using
the finite differences method. **C**, position of hosts and parasites in the
space of latent variables on the first and second dimensions of their respective
subspaces (the results have been clamped to the unit interval). **D**, predicted
interaction weight from the RDPG based on the status of the species pair in the
metaweb.](figures/illustration-part1.png){#fig:illustration1}

In @fig:illustration1, we focus on some statistical checks of the embedding. In
panel **A**, we show that the averaged $L_2$ loss (*i.e.* the sum of squared
errors) between the empirical and reconstructed metaweb decreases when the
number of dimensions (rank) of the subspace increases, with an inflection at 39
dimensions (out of 120 initially) according to the finite differences method. As
discussed by @Runghen2021Exploiting, there is often a trade-off between the
number of dimensions to use (more dimensions are more computationally demanding)
and the quality of the representation. In panel **B**, we show the increase in
cumulative variance explained at each rank, and visualize that using 39 ranks
explains about 70% of the variance in the empirical metaweb. This is a different
information from the $L_2$ loss (which is averaged across interactions), as it
works on the eigenvalues of the embedding, and therefore captures higher-level
features of the network. In panel **C**, we show positions of hosts and
parasites on the first two dimensions of the left and right subspaces. Note that
these values largely skew negative, because the first dimensions capture the
coarse structure of the network: most pairs of species do not interact, and
therefore have negative values. Finally in panel **D**, we show the predicted
weight (*i.e.* the result of the multiplication of the RDGP subspaces at a rank
of 39) as a function of whether the interactions are observed, not-observed, or
unknown due to lack of co-occurrence in the original dataset. This reveals that
the observed interactions have higher predicted weights, although there is some
overlap; the usual approach to identify potential interactions based on this
information would be a thresholding analysis, which is outside the scope of this
manuscript (and is done in the papers cited in this illustration). Because the
values returned from RDPG are not bound to the unit interval, we performed a
clamping of the weights to the unit space, showing a one-inflation in documented
interactions, and a zero-inflation in other species pairs. This last figure
crosses from the statistical into the ecological, by showing that species pairs
with no documented co-occurrence have weights that are not distinguishable from
species pairs with no documented interactions, suggesting that (as befits a
host-parasite model) the ability to interact is a strong predictor of
co-occurrence.

![Ecological analysis of an embedding for a host-parasite metaweb, using Random
Dot Product Graphs. **A**, relationship between the number of parasites and
position along the first axis of the right-subspace for all hosts, showing that
the embedding captures elements of network structure at the species scale.
**B**, weak relationship between the body mass of hosts (in grams) and the
position alongside the same dimension. **C**, weak relationship between body
mass of hosts and parasite richness. **D**, distribution of positions alongside
the same axis for hosts grouped by taxonomic
family.](figures/illustration-part2.png){#fig:illustration2}

The results of @fig:illustration1 show that we can extract an embedding of the
metaweb that captures enough variance to be relevant; specifically, this is true
for both $L_2$ loss (indicating that RDPG is able to capture pairwise processes)
and the cumulative variance explained (indicating that RDPG is able to capture
network-level structure). Therefore, in @fig:illustration2, we relate the values
of latent variables for hosts to different ecologically-relevant data. In panel
**A**, we show that host with a higher value on the first dimension have fewer
parasites. This relates to the body size of hosts in the *PanTHERIA* database
[@Jones2009Pantheria], as shown in panel **B**: interestingly, the position on
the first axis is only weakly correlated to body mass of the host; this matches
well established results showing that body size/mass is not always a direct
predictor of parasite richness in terrestrial mammals [@Morand1998Density], a
result we observe in panel **C**. Finally, in panel **D**, we can see how
different taxonomic families occupy different positions on the first axis, with
*e.g.* Sciuridae being biased towards higher values. These results show how we
can look for ecological informations in the output of network embeddings, which
can further be refined into the selection of predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions between
species at a large spatial scale, typically a scale large enough to be
considered of biogeographic relevance (indeed, many of the examples covered in
the introduction span areas larger than a country, some of them global). But as
@Herbert1965Dune rightfully pointed out, "[y]ou can't draw neat lines around
planet-wide problems"; any inference of a metaweb must therefore contend with
several novel, interwoven, families of problems. In this section, we outline
three that we think are particularly important, and can discuss how they may
addressed with subsequent data analysis or simulations, and how they emerge in
the specific context of using embeddings; some of these issues are related to
the application of these methods at the science-policy interface.

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic point of
view, the chances of finding another area with enough related species (through
phylogenetic relatedness or similarity of functional traits) to make a reliable
inference decreases. This is because transfer requires similarity
(@fig:embedding). A diagnostic for the lack of similar species would likely be
large confidence intervals during estimation of the values in the low-rank
space. In other words, the representation of the original graph is difficult to
transfer to the new problem. Alternatively, if the initial metaweb is too large
(taxonomically), then the resulting embeddings would need to represent
interactions between taxonomic groups that are not present in the new location.
This would lead to a much higher variance in the starting dataset, and to
under-dispersion in the target dataset, resulting in the potential under or over
estimation of the strength of new predicted interactions.
@Llewelyn2022Predicting provided compelling evidence for these situations by
showing that, even at small spatial scales, the transfer of information about
interactions becomes more challenging when areas rich with endemic species are
considered. The lack of well documented metawebs is currently preventing the
development of more concrete guidelines. The question of phylogenetic
relatedness and distribution is notably relevant if the metaweb is assembled in
an area with mostly endemic species (*e.g.* a system that has undergone recent
radiation or that has remained in isolation for a long period of time might not
have an analogous system with which to draw knowledge from), and as with every
predictive algorithm, there is room for the application of our best ecological
judgement. Because this problem relates to distribution of species in the
geographic or phylogenetic space, it can certainly be approached through
assessing the performance of embedding transfer in simulated starting/target
species pools.

@@

# Conclusion: metawebs, predictions, and people

@@

::: {#box-legacy .callout-note}
## Minding legacies shaping ecological datasets {#box-legacy}

@@

**Acknowledgements:** We acknowledge that this study was conducted on land
within the traditional unceded territory of the Saint Lawrence Iroquoian,
Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak nations. TP, TS, DC,
and LP received funding from the Canadian Institute for Ecology & Evolution. FB
is funded by the Institute for Data Valorization (IVADO). TS, SB, and TP are
funded by a donation from the Courtois Foundation. CB was awarded a Mitacs
Elevate Fellowship no. IT12391, in partnership with fRI Research, and also
acknowledges funding from Alberta Innovates and the Forest Resources Improvement
Association of Alberta. M-JF acknowledges funding from NSERC Discovery Grant and
NSERC CRC. RR is funded by New Zealand’s Biological Heritage Ngā Koiora Tuku Iho
National Science Challenge, administered by New Zealand Ministry of Business,
Innovation, and Employment. BM is funded by the NSERC Alexander Graham Bell
Canada Graduate Scholarship and the FRQNT master’s scholarship. LP acknowledges
funding from NSERC Discovery Grant (NSERC RGPIN-2019-05771). TP acknowledges
financial support from the Fondation Courtois, and NSERC through the Discovery
Grants and Discovery Accelerator Supplement programs. MJF is supported by an
NSERC PDF and an RBC Post-Doctoral Fellowship.

>>>>

![The embedding process (**A**) can help to identify links (interactions) that may have been missed within the original community (represented by the orange dashed arrows, **B**). Transfer learning (**D**) allows for the prediction links (interactions) even when novel species (**C**) are included alongside the original community. This is achieved by learning using other relevant predictors (*e.g.* traits) in conjunction with the known interactions to infer latent values (**E**). Ultimately this allows us to predict links (interactions) for species external from the original sample (blue dashed arrows) as well as missing within sample links (**F**). Within this context the predicted (and original) networks as well as the ecological predictors used (green boxes) are products that can be quantified through measurements in the field, whereas the embedded as well as imputed matrices (purple box) are representative of a decomposition of the interaction matrices onto the embedding space](figures/conceptual_2.png){#fig-embedding}

@@

Graph (or network) embedding (@fig-embedding) is a family of machine learning techniques, whose main task is to learn a mapping function from a discrete graph to a continuous domain [@Arsov2019Network; @Chami2022Machine]. Their main goal is to learn a low dimensional vector representation of the graph (embeddings), such that its key properties (*e.g.* local or global structures) are retained in the embedding space [@Yan2005Graph]. The embedding space may, but will not necessarily, have lower dimensionality than the graph. Ecological networks are promising candidates for the routine application of embeddings, as they tend to possess a shared structural backbone [see *e.g.* @BramonMora2018Identifying], which hints at structural invariants in empirical data. Assuming that these structural invariants are common enough, they would dominate the structure of networks, and therefore be adequately captured by the first (lower) dimensions of an embedding, without the need to measure derived aspects of their structure (*e.g.* motifs, paths, modularity, ...).

@@

Food webs are inherently low-dimensional objects, and can be adequately represented with less than ten dimensions [@Braga2021Phylogenetic; @Eklof2013Dimensionality; @Braga2019Spatial]. Simulation results by @Botella2022Appraisal suggested that there is no dominant method to identify architectural similarities between networks: multiple approaches need to be tested and compared to the network descriptor of interest on a problem-specific basis. This matches previous results on graph embedding, wherein different embedding algorithms yield different network embeddings [@Goyal2018Graph], calling for a careful selection of the problem-specific approach to use. In @tbl-methods, we present a selection of common graph and node embedding methods, alongside examples of their use to predict interactions or statistical associations between species. These methods rely largely on linear algebra or pseudo-random walks on graphs. All forms of embeddings presented in @tbl-methods share the common property of summarizing their objects into (sets of) dense feature vectors, that capture the overall network structure, pairwise information on nodes, and emergent aspects of the network, in a compressed way (*i.e.* with some information loss, as we later discuss in the illustration). Node embeddings tend to focus on maintaining pairwise relationships (*i.e.* species interactions), while graph embeddings focus on maintaining the network structure (*i.e.* emergent properties). Nevertheless, some graph embedding techniques [like RDPG, see *e.g.* @Wu2021Maximum] will provide high-quality node-level embeddings while also preserving network structure.

::: {#box-gnn .callout-note}
## Box 1 - Graph Neural Networks

One prominent family of approaches we do not discuss in the present manuscript is Graph Neural Networks [GNN; @Zhou2020Graph]. GNN are, in a sense, a method to embed a graph into a dense subspace, but belong to the family of deep learning methods, which has its own set of practices [see *e.g.* @Goodfellow2016Deep]. An important issue with methods based on deep learning is that, because their parameter space is immense, the sample size of the data fed into them must be similarly large (typically thousands of instances). This is a requirement for the model to converge correctly during training, but this assumption is unlikely to be met given the size of datasets currently available for metawebs (or single time/location species interaction networks). This data volume requirement is mostly absent from the techniques we list below. Furthermore, GNN still have some challenges related to their shallow structure, and concerns related to scalability [see @Gupta2021Graph for a review], which are mostly absent from the methods listed in @tbl-methods. Assuming that the uptake of next-generation biomonitoring techniques does indeed deliver larger datasets on species interactions [@Bohan2017Nextgeneration], there is nevertheless the potential for GNN to become an applicable embedding/predictive technique in the coming years.
:::

Graph embeddings *can* serve as a dimensionality reduction method. For example, RDPG [@Strydom2022Food] and t-SVD [truncated Singular Value Decomposition; @Poisot2021Imputing] typically embed networks using fewer dimensions than the original network [the original network has as many dimensions as species, and as many informative dimensions as trophically unique species; @Strydom2021Svd]. However, this is not necessarily the case -- indeed, one may perform a PCA (a special case of SVD) to project the raw data into a subspace that improves the efficacy of t-SNE [t-distributed stochastic neighbor embedding; @Maaten2009Learning]. There are many dimensionality reductions [@Anowar2021Conceptual] that can be applied to an embedded network should the need for dimensionality reduction (for example for data visualization) arise. In brief, many graph embeddings *can* serve as dimensionality reduction steps, but not all do, neither do all dimensionality reduction methods provide adequate graph embedding capacities. In the next section (and @fig-embedding), we show how the amount of dimensionality reduction can affect the quality of the embedding.

| Method        | Object          | Technique                        | Reference                | Application                                                                                                       |
|-------------|-------------|-------------|-------------|--------------------|
| tSNE          | nodes           | statistical divergence           | @Hinton2002Stochastic    | [@Cieslak2020Tdistributed, species-environment responses $^a$] [@Gibb2021Data, host-virus network representation] |
| LINE          | nodes           | stochastic gradient descent      | @Tang2015Line            |                                                                                                                   |
| SDNE          | nodes           | gradient descent                 | @Wang2016Structural      |                                                                                                                   |
| node2vec      | nodes           | stochastic gradient descent      | @Grover2016Node2vec      |                                                                                                                   |
| HARP          | nodes           | meta-strategy                    | @Chen2017Harp            |                                                                                                                   |
| DMSE          | joint nodes     | deep neural network              | @Chen2017Deep            | [@Chen2017Deep, species-environment interactions $^b$]                                                            |
| graph2vec     | sub-graph       | skipgram network                 | @Narayanan2017Graph2vec  |                                                                                                                   |
| RDPG          | graph           | SVD                              | @Young2007Random         | [@DallaRiva2016Exploring, trophic interactions] [@Poisot2021Imputing, host-virus network prediction]              |
| GLEE          | graph           | Laplacian eigenmap               | @Torres2020Glee          |                                                                                                                   |
| DeepWalk      | graph           | stochastic gradient descent      | @Perozzi2014Deepwalk     | [@Wardeh2021Predicting, host-virus interactions]                                                                  |
| GraphKKE      | graph           | stochastic differential equation | @Melnyk2020Graphkke      | [@Melnyk2020Graphkke, microbiome species associations $^a$]                                                       |
| FastEmbed     | graph           | eigen decomposition              | @Ramasamy2015Compressive |                                                                                                                   |
| PCA           | graph           | eigen decomposition              | @Surendran2013Graph      | [@Strydom2021Roadmap, host-parasite interactions]                                                                 |
| Joint methods | multiple graphs | multiple strategies              | @Wang2021Joint           |                                                                                                                   |

: Overview of some common graph embedding approaches, by type of embedded objects, alongside examples of their use in the prediction of species interactions. These methods have not yet been routinely used to predict species interactions; most examples that we identified were either statistical associations, or analogues to joint species distribution models. $^a$: application is concerned with *statistical* interactions, which are not necessarilly direct biotic interactions; $^b$:application is concerned with joint-SDM-like approach, which is also very close to statistical associations as opposed to direct biotic interactions. Given the need to evaluate different methods on a problem-specific basis, the fact that a lot of methods have not been used on network problems is an opportunity for benchmarking and method development. Note that the row for PCA also applies to kernel/probabilistic PCA, which are variations on the more general method of SVD. Note further that tSNE has been included because it is frequently used to embed graphs, including of species associations/interactions, despite not being strictly speaking, a graph embedding technique [see *e.g.* @Chami2022Machine]. {#tbl-methods}

The popularity of graph embedding techniques in machine learning is more than the search for structural invariants: graphs are discrete objects, and machine learning techniques tend to handle continuous data better. Bringing a sparse graph into a continuous, dense vector space [@Xu2021Understanding] opens up a broader variety of predictive algorithms, notably of the sort that are able to predict events as probabilities [@Murphy2022Probabilistic]. Furthermore, the projection of the graph itself is a representation that can be learned; @Runghen2021Exploiting, for example, used a neural network to learn the embedding of a network in which not all interactions were known, based on the nodes' metadata. This example has many parallels in ecology (see @fig-embedding **C**), in which node metadata can be represented by phylogeny, abundance, or functional traits. Using phylogeny as a source of information assumes (or strives to capture) the action of evolutionary processes on network structure, which at least for food webs have been well documented [@Braga2021Phylogenetic; @DallaRiva2016Exploring; @Eklof2016Phylogenetic; @Stouffer2007Evidence; @Stouffer2012Evolutionary]; similarly, the use of functional traits assumes that interactions can be inferred from the knowledge of trait-matching rules, which is similarly well supported in the empirical literature [@Bartomeus2013Understanding; @Bartomeus2016Common; @Goebel2023Body; @Gravel2013Inferring]. Relating this information to an embedding rather than a list of network measures would allow to capture their effect on the more fundamental aspects of network structure; conversely, the absence of a phylogenetic or functional signal may suggest that evolutionary/trait processes are not strong drivers of network structure, therefore opening a new way to perform hypothesis testing.

@@

{{< embed notebooks/SupplementaryMaterial.ipynb#fig-illustration-1 >}}

In @fig-illustration-1, we focus on some statistical checks of the embedding. In panel **A**, we show that the averaged $L_2$ loss (*i.e.* the mean of squared errors) between the empirical and reconstructed metaweb decreases when the number of dimensions (rank) of the subspace increases, with an inflection at 39 dimensions (out of 120 initially) according to the finite differences method. As discussed by @Runghen2021Exploiting, there is often a trade-off between the number of dimensions to use (more dimensions are more computationally demanding) and the quality of the representation. In panel **B**, we show the increase in cumulative variance explained at each rank, and visualize that using 39 ranks explains about 70% of the variance in the empirical metaweb. This is a different information from the $L_2$ loss (which is averaged across interactions), as it works on the eigenvalues of the embedding, and therefore captures higher-level features of the network. In panel **C**, we show positions of hosts and parasites on the first two dimensions of the left and right subspaces. Note that these values largely skew negative, because the first dimensions capture the coarse structure of the network: most pairs of species do not interact, and therefore have negative values. Finally in panel **D**, we show the predicted weight (*i.e.* the result of the multiplication of the RDGP subspaces at a rank of 39) as a function of whether the interactions are observed, not-observed, or unknown due to lack of co-occurrence in the original dataset. This reveals that the observed interactions have higher predicted weights, although there is some overlap; the usual approach to identify potential interactions based on this information would be a thresholding analysis, which is outside the scope of this manuscript (and is done in the papers cited in this illustration). Because the values returned from RDPG are not bound to the unit interval, we performed a clamping of the weights to the unit space, showing a one-inflation in documented interactions, and a zero-inflation in other species pairs. This last figure crosses from the statistical into the ecological, by showing that species pairs with no documented co-occurrence have weights that are not distinguishable from species pairs with no documented interactions, suggesting that (as befits a host-parasite model) the ability to interact is a strong predictor of co-occurrence.

{{< embed notebooks/SupplementaryMaterial.ipynb#fig-illustration-2 >}}

The results of @fig-illustration-1 show that we can extract an embedding of the metaweb that captures enough variance to be relevant; specifically, this is true for both $L_2$ loss (indicating that RDPG is able to capture pairwise processes) and the cumulative variance explained (indicating that RDPG is able to capture network-level structure). Therefore, in @fig-illustration-2, we relate the values of latent variables for hosts to different ecologically-relevant data. In panel **A**, we show that host with a higher value on the first dimension have fewer parasites. This relates to the body size of hosts in the *PanTHERIA* database [@Jones2009Pantheria], as shown in panel **B**: interestingly, the position on the first axis is only weakly correlated to body mass of the host; this matches well established results showing that body size/mass is not always a direct predictor of parasite richness in terrestrial mammals [@Morand1998Density], a result we observe in panel **C**. Finally, in panel **D**, we can see how different taxonomic families occupy different positions on the first axis, with *e.g.* Sciuridae being biased towards higher values. These results show how we can look for ecological informations in the output of network embeddings, which can further be refined into the selection of predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions between species at a large spatial scale, typically a scale large enough to be considered of biogeographic relevance (indeed, many of the examples covered in the introduction span areas larger than a country, some of them global). But as @Herbert1965Dune rightfully pointed out, "\[y\]ou can't draw neat lines around planet-wide problems"; any inference of a metaweb must therefore contend with several novel, interwoven, families of problems. In this section, we outline three that we think are particularly important, and can discuss how they may addressed with subsequent data analysis or simulations, and how they emerge in the specific context of using embeddings; some of these issues are related to the application of these methods at the science-policy interface. Adressing these considerations as part of the methodological discussion is particularly important, as the construction of metawebs can perpetuate legacies of biases in data (Box 2).

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic point of view, the chances of finding another area with enough related species (through phylogenetic relatedness or similarity of functional traits) to make a reliable inference decreases. This is because transfer requires similarity (@fig-embedding). A diagnostic for the lack of similar species would likely be large confidence intervals during estimation of the values in the low-rank space. In other words, the representation of the original graph is difficult to transfer to the new problem. Alternatively, if the initial metaweb is too large (taxonomically), then the resulting embeddings would need to represent interactions between taxonomic groups that are not present in the new location. This would lead to a much higher variance in the starting dataset, and to under-dispersion in the target dataset, resulting in the potential under or over estimation of the strength of new predicted interactions. @Llewelyn2022Predicting provided compelling evidence for these situations by showing that, even at small spatial scales, the transfer of information about interactions becomes more challenging when areas rich with endemic species are considered. The lack of well documented metawebs is currently preventing the development of more concrete guidelines. The question of phylogenetic relatedness and distribution is notably relevant if the metaweb is assembled in an area with mostly endemic species (*e.g.* a system that has undergone recent radiation or that has remained in isolation for a long period of time might not have an analogous system with which to draw knowledge from), and as with every predictive algorithm, there is room for the application of our best ecological judgement. Because this problem relates to distribution of species in the geographic or phylogenetic space, it can certainly be approached through assessing the performance of embedding transfer in simulated starting/target species pools.

@@

# Conclusion

@@

::: {#sec-box-legacy .callout-note}
## Box 2 - Minding legacies shaping ecological datasets

@@

**Acknowledgements:** We acknowledge that this study was conducted on land within the traditional unceded territory of the Saint Lawrence Iroquoian, Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak nations. TP, TS, DC, and LP received funding from the Canadian Institute for Ecology & Evolution. FB is funded by the Institute for Data Valorization (IVADO). TS, SB, and TP are funded by a donation from the Courtois Foundation. CB was awarded a Mitacs Elevate Fellowship no. IT12391, in partnership with fRI Research, and also acknowledges funding from Alberta Innovates and the Forest Resources Improvement Association of Alberta. M-JF acknowledges funding from NSERC Discovery Grant and NSERC CRC. RR is funded by New Zealand's Biological Heritage Ngā Koiora Tuku Iho National Science Challenge, administered by New Zealand Ministry of Business, Innovation, and Employment. BM is funded by the NSERC Alexander Graham Bell Canada Graduate Scholarship and the FRQNT master's scholarship. LP acknowledges funding from NSERC Discovery Grant (NSERC RGPIN-2019-05771). TP acknowledges financial support from the Fondation Courtois, and NSERC through the Discovery Grants and Discovery Accelerator Supplement programs. MJF is supported by an NSERC PDF and an RBC Post-Doctoral Fellowship.
